Understanding the Landscape of Childlessness in the U.S.

### 2024-03-14 outputs/paper/paper.qmd
Graphs moved to the data section

---
title: "Understanding the Landscape of Childlessness in the U.S.: A Data-Driven Exploration (1972-2022)"
author: 
  - Mehrnoush Mohammadi 
  - Quang Mai
thanks: "Code and data are available at: https://github.com/mehrnoush68/childlessness-in-america.git"
date: today
date-format: long
abstract: "This paper investigates U.S. childlessness trends (1972-2022) using data collected from the General Social Survey (GSS). The research covers several factors such as age, gender, education, social class, health conditions, marital status, and number of children. The analysis reveals rising childlessness across demographics, potentially linked to economic factors, education, and social norms. We explore the social and economic implications of this trend and emphasize the need for further research on childlessness in modern America."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

#### Load data ####
table_nochild_age <- read_csv(file = "../data/clean_data_age.csv",show_col_types = FALSE)

table_nochild_class <- read_csv(file = "../data/clean_data_class.csv",show_col_types = FALSE)

table_nochild_degree <- read_csv(file = "../data/clean_data_degree.csv",show_col_types = FALSE)

table_nochild_sex <- read_csv(file = "../data/clean_data_sex.csv",show_col_types = FALSE)

table_nochild_marital <- read_csv(file = "../data/clean_data_marital.csv",show_col_types = FALSE)

table_nochild_health <- read_csv(file = "../data/clean_data_health.csv",show_col_types = FALSE)

@@

# Introduction {#sec-intro}
The American family landscape is undergoing a significant transformation, marked by a growing trend of childlessness. Traditionally viewed as a societal norm, having children now becomes a choice [@voluntaryChild]. This paper looks into this phenomenon by examining various datasets on factors such as age, gender, education, socioeconomic class, health conditions and marital status, and how they affect childbearing rates over the past 50 years.

Analyzing these factors offers a window into understanding not only the prevalence of childlessness within different segments of the population but also the potential drivers behind this trend. Existing research suggests a complex interplay of factors, including economic considerations, rising educational attainment, changing social norms related to career and family life, and individual preferences [@voluntaryChild].

This paper aims to unpack these factors and explore their impact on the rise of childlessness in America. By analyzing the data, we seek to answer crucial questions: What are some important factors contributing to this trend? How does childlessness manifest differently across demographic groups? Finally, what are the potential social, economic, and individual consequences of this new family structure in the United States? Understanding these dynamics is crucial in having an informed policy discussion and navigating the implications of childlessness in a rapidly changing world.

Structurally, in [Introduction], our paper starts with a brief dissection of the nature of childlessness in America and the contributing factors. Subsequently, in [Data] and [Result], we talk about the nature of the data obtained and analyze the results garnered from the data with suitable tables and charts. Next, [Discussion] provides further insights and future areas of study. Finally, [Conclusion] concludes and summarizes our main findings. To complete the paper, [Appendix] clarifies how each variable within each dataset is measured.

@@

# Data {#sec-data}
This paper uses a multifaceted approach, drawing on various datasets to understand the complexities of childlessness in the United States. One primary source is the General Social Survey (GSS), a long-running and highly respected survey conducted by the National Opinion Research Center (NORC) at the University of Chicago [@gss], from 1972 to 2022.

@@

* Children: This variable captures the number of children ever born alive to the respondent, providing a direct measure of childlessness (having no children).
* Year: This variable allows us to analyze trends in childlessness over time, encompassing the chosen time frame for the survey.
* Age: This variable, categorized into relevant age groups (e.g., 18-34, 35-49, etc.), enables us to understand how childlessness manifests across different age demographics.

@@

### Exploring Socioeconomic Class and Educational Attainment
To gain a deeper understanding of the factors influencing these trends, we look into additional data sets from NORC (@gss) examining the relationship between childlessness and socioeconomic class @tbl-nochild-class and educational attainment @tbl-nochild-degree.

@@

* Socioeconomic Class: Data on socioeconomic class, often measured by income or occupation, will be utilized to understand the impact of economic factors on childbearing choices.
* Educational Attainment: Data on educational attainment, such as the highest degree obtained, will be incorporated to explore the relationship between educational level and childbearing.

@@

# Print the table with formatting
kable(table_nochild_class_subset, 
      caption = "Childless Percentage of Population by Class Indentification and Selected Years", 
      align = "c",
      col.names = random_years,
      row.names = FALSE) |>
  kable_styling(latex_options = c("striped", "hold_position"), bootstrap_options = "striped", full_width = FALSE)

@@

@@

```{r}
#| echo: false
#| include: true
#| warning: false
#| label: tbl-nochild-degree

@@

### Exploring Gender Differences
It is crucial to acknowledge potential gender disparities in childbearing experiences. A separate analysis examining childlessness by gender (presented in @tbl-nochild-sex) aims to provide a comprehensive understanding by considering the unique experiences of men and women regarding childlessness. Data on gender is included from NORC [@gss] to allow for this investigation.

@@

### Exploring Health Conditions and Marital Status

In addition, additional data sets from NORC [@gss] were also leveraged to understand the context of childlessness in America from 1972-2022. @tbl-nochild-health and @tbl-nochild-marital present an interesting connection between childlessness and health conditions and martial status.

@@

@@

@fig-nochild-age shows a consistent rise in childlessness across all age groups in the United States from 1972 to 2022.

ggplot(data_long_age, aes(x = as.integer(year), y = percentage, color = Age)) +
  geom_line() +
  geom_smooth(method = "loess", se = FALSE, linetype = "dashed", size = 1.2) +  # Add running average trend line
  labs(x = "Year", y = "Percentage of Population", title = "Childless Percentage of Population over Time by Age Group", color = "Age Group") +
  theme_minimal() +
  theme(text = element_text(size = 10))

It is important to acknowledge, however, that the data series has gaps for several years within this timeframe (1979, 1981, 1992, 1995, 1997, 1999, 2001, 2003, 2005, 2007, 2009, 2011, 2013, 2015, 2017, and 2019). These missing data points can potentially limit the generalization of the observed trends and hinder a comprehensive understanding of childlessness over time.
Despite these limitations, the available data reveals a steady increase in the percentage of individuals without children across all age groups. Notably, the most significant rise is observed among younger adults (18-34 and 35-49). As an example, the percentage of childless individuals in the 18-34 age group increased by approximately 42.70% from 1972 to 2022.

# Plot the data
ggplot(data_long_class, aes(x = Year, y = Percentage, color = `Class Identification`)) +
  geom_line() +
  geom_smooth(method = "loess", se = FALSE, linetype = "dashed", size = 1.2) +  # Add running average trend line
  labs(x = "Year", y = "Percentage of Population", title = "Childless Percentage of Population over Time by Class", color = "Breakdown Category") +
  theme_minimal() +
  theme(text = element_text(size = 10))

@@

The data reveals a gradual increase in childlessness for both genders over the past five decades. However, females consistently exhibit a lower percentage of childlessness compared to males throughout the entire period. While the gap appears to be slightly narrowing over time, females still have a lower percentage by around 8 percentage points in 2022 (approximately 28.32% for females and 36.09% for males).

# Plot the data
ggplot(data_long_marital, aes(x = Year, y = Percentage, color = `Marital status`)) +
  geom_line() +
  geom_smooth(method = "loess", se = FALSE, linetype = "dashed", size = 1.2) +  # Add running average trend line
  labs(x = "Year", y = "Percentage of Population", title = "Childless Percentage of Population over Time by Marital Status", color = "Marital status") +
  theme_minimal() +
  theme(text = element_text(size = 10))

```{r}
#| echo: false
#| include: true
#| warning: false
#| label: fig-nochild-avg-marital
#| fig-cap: "Average Childless Percentage of Population over Time by Marital Status"

# Group the data by Year and calculate the average of Percentage for each year
avg_data <- data_long_marital |>
  group_by(Year) |>
  summarise(Sum_Percentage = sum(Percentage, na.rm = TRUE)) |>
  mutate(avg_Percentage = Sum_Percentage / 5)

# Print the average data
# print(avg_data)

# Plot the data
ggplot(avg_data, aes(x = Year, y = avg_Percentage)) +
  geom_line() +
  geom_point() +
  labs(x = "Year", y = "Average Percentage of Population", title = "Average Childless Percentage of Population over Time by Marital Status") +
  theme_minimal()

```
```{r}
#| echo: false
#| include: true
#| warning: false
#| label: fig-nochild-avg-class
#| fig-cap: "Average Percentage Childless over Time by Class"

# Group the data by Year and calculate the average of Percentage for each year
avg_data <- data_long_class |>
  group_by(Year) |>
  summarise(Sum_Percentage = sum(Percentage, na.rm = TRUE)) |>
  mutate(avg_Percentage = Sum_Percentage / 4)

# Print the average data
# print(avg_data)

# Plot the data
ggplot(avg_data, aes(x = Year, y = avg_Percentage)) +
  geom_line() +
  geom_point() +
  labs(x = "Year", y = "Average Percentage of Population", title = "Average Percentage Childless over Time by Class") +
  theme_minimal()

@@

# Group the data by Year and calculate the average of Percentage for each year
avg_data <- data_long_sex |>
  group_by(Year) |>
  summarise(Sum_Percentage = sum(Percentage, na.rm = TRUE)) |>
  mutate(avg_Percentage = Sum_Percentage / 2)

# Print the average data
# print(avg_data)

# Plot the data
ggplot(avg_data, aes(x = Year, y = avg_Percentage)) +
  geom_line() +
  geom_point() +
  labs(x = "Year", y = "Average Percentage of Population", title = "Average Percentage Childless over Time by Gender") +
  theme_minimal()

@@

# Group the data by Year and calculate the average of Percentage for each year
avg_data <- data_long_degree |>
  group_by(Year) |>
  summarise(Sum_Percentage = sum(Percentage, na.rm = TRUE)) |>
  mutate(avg_Percentage = Sum_Percentage / 3)

# Print the average data
# print(avg_data)

# Plot the data
ggplot(avg_data, aes(x = Year, y = avg_Percentage)) +
  geom_line() +
  geom_point() +
  labs(x = "Year", y = "Average Percentage of Population", title = "Average Percentage Childless over Time by Degree") +
  theme_minimal()

@@

@@

## Rising Childlessness: Patterns and Trends
The findings presented in this paper reveal a consistent and concerning trend of rising childlessness across various demographics in the United States. As Figure 1 illustrates, the percentage of individuals without children has steadily increased across all age groups over the past 50 years, with the most significant rise observed among younger individuals (18-34 and 35-49).

>>>>

---
title: "Understanding the Landscape of Childlessness in the U.S."
Subtitle: "An analysis of General Social Survey data (1972-2022)"
author: 
  - Mehrnoush Mohammadi 
  - Quang Mai
thanks: "Code and data are available at: https://github.com/mehrnoush68/childlessness-in-america.git"
date: today
date-format: long
abstract: "We investigate U.S. childlessness trends between 1972 and 2022 using data from the General Social Survey. The research focuses on the relationship between childlessness and factors such as age, gender, education, social class, health conditions, marital status, and number of children. We find rising childlessness across demographics, potentially linked to economic factors, education, and social norms. We explore the social and economic implications of this trend and emphasize the need for further research on childlessness."
format: pdf
number-sections: true
toc: true
bibliography: references.bib
---

@@

#### Load data ####
table_nochild_age <- read_csv(file = "../data/clean_data_age.csv",show_col_types = FALSE)
table_nochild_class <- read_csv(file = "../data/clean_data_class.csv",show_col_types = FALSE)
table_nochild_degree <- read_csv(file = "../data/clean_data_degree.csv",show_col_types = FALSE)
table_nochild_sex <- read_csv(file = "../data/clean_data_sex.csv",show_col_types = FALSE)
table_nochild_marital <- read_csv(file = "../data/clean_data_marital.csv",show_col_types = FALSE)
table_nochild_health <- read_csv(file = "../data/clean_data_health.csv",show_col_types = FALSE)

@@

# Introduction {#sec-intro}
The American family landscape is undergoing a significant transformation, marked by a growing trend of childlessness. Traditionally viewed as a societal norm, having children now becomes a choice [@voluntaryChild]. We examine factors such as age, gender, education, socioeconomic class, health conditions and marital status, and how they affect childbearing rates over the past 50 years.

This offers a window into understanding not only the prevalence of childlessness within different segments of the population but also the potential drivers behind this trend. Existing research suggests a complex interplay of factors, including economic considerations, rising educational attainment, changing social norms related to career and family life, and individual preferences [@voluntaryChild].

We explore the impact of these factors on the rise of childlessness in America. By analyzing the data, we seek to answer crucial questions: What are some important factors contributing to this trend? How does childlessness manifest differently across demographic groups? Finally, what are the potential social, economic, and individual consequences of this new family structure in the United States? Understanding these dynamics is crucial in having an informed policy discussion and navigating the implications of childlessness in a rapidly changing world.

Structurally, in [Introduction], our paper starts with a brief dissection of the nature of childlessness in America and the contributing factors. Subsequently, in [Data] and [Result], we talk about the nature of the data obtained and analyze the results garnered from the data with suitable tables and charts. Next, [Discussion] provides further insights and future areas of study. Finally, [Conclusion] summarizes our main findings. To complete the paper, [Appendix] clarifies how each variable within each dataset is measured.

@@

# Data {#sec-data}
This paper uses a multifaceted approach, drawing on various datasets to understand the complexities of childlessness in the United States. Our primary source is the General Social Survey (GSS), a long-running and highly respected survey conducted by the National Opinion Research Center (NORC) at the University of Chicago [@gss], from 1972 to 2022.

@@

* **Children:** This variable captures the number of children ever born alive to the respondent, providing a direct measure of childlessness (having no children).
* **Year:** This variable allows us to analyze trends in childlessness over time, encompassing the chosen time frame for the survey.
* **Age:** This variable, categorized into relevant age groups (e.g., 18-34, 35-49, etc.), enables us to understand how childlessness manifests across different age demographics.

@@

@fig-nochild-age shows a consistent rise in childlessness across all age groups in the United States from 1972 to 2022. It is important to acknowledge, however, that the data series has gaps for several years within this timeframe (1979, 1981, 1992, 1995, 1997, 1999, 2001, 2003, 2005, 2007, 2009, 2011, 2013, 2015, 2017, and 2019). These missing data points can potentially limit the generalization of the observed trends and hinder a comprehensive understanding of childlessness over time.
Despite these limitations, the available data reveals a steady increase in the percentage of individuals without children across all age groups. Notably, the most significant rise is observed among younger adults (18-34 and 35-49). As an example, the percentage of childless individuals in the 18-34 age group increased by approximately 42.70% from 1972 to 2022.

ggplot(data_long_age, aes(x = as.integer(year), y = percentage, color = Age)) +
  geom_point() +
  geom_smooth(method = "loess", se = FALSE, linetype = "solid", size = 1.2) +  # Add running average trend line
  labs(x = "Year", y = "Percentage of Population", title = "Childless Percentage of Population over Time by Age Group", color = "Age Group") +
  theme_minimal() +
  theme(text = element_text(size = 10))

## Exploring Socioeconomic Class and Educational Attainment
To gain a deeper understanding of the factors influencing these trends, we look into additional data sets from NORC [@gss] examining the relationship between childlessness and socioeconomic class @tbl-nochild-class and educational attainment @tbl-nochild-degree.

@@

* **Socioeconomic Class:** Data on socioeconomic class, often measured by income or occupation, will be utilized to understand the impact of economic factors on childbearing choices.
* **Educational Attainment:** Data on educational attainment, such as the highest degree obtained, will be incorporated to explore the relationship between educational level and childbearing.

@@

# Print the table with formatting
kable(table_nochild_class_subset, 
      caption = "Childless Percentage of Population by Class Identification and Selected Years", 
      align = "c",
      col.names = random_years,
      row.names = FALSE) |>
  kable_styling(latex_options = c("striped", "hold_position"), bootstrap_options = "striped", full_width = FALSE)

@@

# Plot the data
ggplot(data_long_class, aes(x = Year, y = Percentage, color = `Class Identification`)) +
  geom_point() +
  geom_smooth(method = "loess", se = FALSE, linetype = "solid", size = 1.2) +  # Add running average trend line
  labs(x = "Year", y = "Percentage of Population", title = "Childless Percentage of Population over Time by Class", color = "Breakdown Category") +
  theme_minimal() +
  theme(text = element_text(size = 10))

@@

\newpage
```{r}
#| echo: false
#| include: true
#| warning: false
#| label: tbl-nochild-degree

@@

## Exploring Gender Differences
It is crucial to acknowledge potential gender disparities in childbearing experiences. A separate analysis examining childlessness by gender (presented in @tbl-nochild-sex) aims to provide a comprehensive understanding by considering the unique experiences of men and women regarding childlessness. Data on gender is included from NORC [@gss] to allow for this investigation.

@@

```
The data reveals a gradual increase in childlessness for both genders over the past five decades. However, females consistently exhibit a lower percentage of childlessness compared to males throughout the entire period. While the gap appears to be slightly narrowing over time, females still have a lower percentage by around 8 percentage points in 2022 (approximately 28.32% for females and 36.09% for males).

## Exploring Health Conditions and Marital Status

In addition, additional data sets from NORC [@gss] were also leveraged to understand the context of childlessness in America from 1972-2022. @tbl-nochild-health and @tbl-nochild-marital present an interesting connection between childlessness and health conditions and marital status.

@@

# Plot the data
ggplot(data_long_marital, aes(x = Year, y = Percentage, color = `Marital status`)) +
  geom_point() +
  geom_smooth(method = "loess", se = FALSE, linetype = "solid", size = 1.2) +  # Add running average trend line
  labs(x = "Year", y = "Percentage of Population", title = "Childless Percentage of Population over Time by Marital Status", color = "Marital status") +
  theme_minimal() +
  theme(text = element_text(size = 10))

@@

```{r}
#| echo: false
#| include: true
#| warning: false
#| label: fig-nochild-avg-marital
#| fig-cap: "Average Percentage Childless over Time by Marital Status"

# Group the data by Year and calculate the average of Percentage for each year
avg_data_marital <- data_long_marital |>
  group_by(Year) |>
  summarise(Sum_Percentage = sum(Percentage, na.rm = TRUE)) |>
  mutate(avg_Percentage = Sum_Percentage / 5)

# Print the average data
# print(avg_data_marital)

# Plot the data
ggplot(avg_data_marital, aes(x = Year, y = avg_Percentage)) +
  geom_line() +
  geom_point() +
  labs(x = "Year", y = "Average Percentage of Population", title = "Average Childless Percentage of Population over Time by Marital Status") +
  theme_minimal() +
  theme(text = element_text(size = 9))

@@

```{r}
#| echo: false
#| include: true
#| warning: false
#| label: fig-nochild-avg-class
#| fig-cap: "Average Percentage Childless over Time by Class"

# Group the data by Year and calculate the average of Percentage for each year
avg_data_class <- data_long_class |>
  group_by(Year) |>
  summarise(Sum_Percentage = sum(Percentage, na.rm = TRUE)) |>
  mutate(avg_Percentage = Sum_Percentage / 4)

# Print the average data
# print(avg_data_class)

# Plot the data
ggplot(avg_data_class, aes(x = Year, y = avg_Percentage)) +
  geom_line() +
  geom_point() +
  labs(x = "Year", y = "Average Percentage of Population", title = "Average Percentage Childless over Time by Class") +
  theme_minimal() +
  theme(text = element_text(size = 9))

@@

# Group the data by Year and calculate the average of Percentage for each year
avg_data_sex <- data_long_sex |>
  group_by(Year) |>
  summarise(Sum_Percentage = sum(Percentage, na.rm = TRUE)) |>
  mutate(avg_Percentage = Sum_Percentage / 2)

# Print the average data
# print(avg_data_sex)

# Plot the data
ggplot(avg_data_sex, aes(x = Year, y = avg_Percentage)) +
  geom_line() +
  geom_point() +
  labs(x = "Year", y = "Average Percentage of Population", title = "Average Percentage Childless over Time by Gender") +
  theme_minimal() +
  theme(text = element_text(size = 9))

@@

# Group the data by Year and calculate the average of Percentage for each year
avg_data_degree <- data_long_degree |>
  group_by(Year) |>
  summarise(Sum_Percentage = sum(Percentage, na.rm = TRUE)) |>
  mutate(avg_Percentage = Sum_Percentage / 3)

# Print the average data
# print(avg_data_degree)

# Plot the data
ggplot(avg_data_degree, aes(x = Year, y = avg_Percentage)) +
  geom_line() +
  geom_point() +
  labs(x = "Year", y = "Average Percentage of Population", title = "Average Percentage Childless over Time by Degree") +
  theme_minimal() +
  theme(text = element_text(size = 9))

@@

```{r}
#| echo: false
#| include: true
#| warning: false
#| label: fig-nochild-avg-degree-class
#| fig-cap: "Average Percentage Childless over Time by Degree and Class"

# Merge the two datasets based on the Year column
merged_data <- merge(avg_data_degree, avg_data_class, by = "Year", all = TRUE, suffixes = c("_degree", "_class"))

# Calculate the sum of avg_Percentage for each year
merged_data <- merged_data |>
  mutate(sum_Percentage = (coalesce(avg_Percentage_degree, 0) + coalesce(avg_Percentage_class, 0))/2)

# Print the merged and summed data
# print(merged_data)

# Plot the merged and summed data
ggplot(merged_data, aes(x = Year, y = sum_Percentage)) +
  geom_line() +
  geom_point() +
  labs(x = "Year", y = "Sum of Average Percentage of Population", title = "Sum of Average Percentage Childless over Time by Degree and Class") +
  theme_minimal() +
  theme(text = element_text(size = 9))
```

@@

## Rising Childlessness: Patterns and Trends
The findings presented in this paper reveal a consistent and concerning trend of rising childlessness across various demographics in the United States. As @fig-nochild-age illustrates, the percentage of individuals without children has steadily increased across all age groups over the past 50 years, with the most significant rise observed among younger individuals (18-34 and 35-49).
